---
title: "Longitudinal Data Analysis Results"
author: "Eric Scott"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## target knits Rmds in their own session, so load libraries here.
## source("../packages.R")
library(targets)
library(lme4)
library(car)
library(tidyverse)
library(emmeans)
```

```{r}
withr::with_dir(here::here(), {
  tar_load(c(
    k_ht,
    b_ht,
    o_ht,
    longdata
  ))
})
```

# Analysis

## Kale

Model:

```{r}
formula(k_ht)
```


```{r}
Anova(k_ht)
```

- days = significant growth (slope of height over time) 
- days:hist_treat:curr_treat = hard to interpret, especially since no main effects

```{r}
ggplot(my_augment(k_ht), aes(x = days, color = hist_treat, linetype = curr_treat)) + 
  geom_line(aes(y = .fitted)) +
  labs(y = "log(fitted height)")
```

Since the difference in the slopes is the focus, I can plot just the slopes for each line, with confidence intervals.  95% CIs are typical, but here I'm plotting 84% CIs which overlap about 95% of the time for random samples from the same population (i.e. null hypothesis) (Payton et al. 2003).

```{r}
k_slopes <- emtrends(k_ht, ~ hist_treat*curr_treat, var = "days", level = 0.84)
k_slope_df <- 
  k_slopes %>% 
  as_tibble() %>% 
  # back-transform linear slope to exponential growth rate?
  mutate(across(c(days.trend, lower.CL, upper.CL), ~exp(.x)))

ggplot(k_slope_df) +
  geom_pointrange(aes(
    x = curr_treat,
    color = hist_treat,
    y = days.trend,
    ymin = lower.CL,
    ymax = upper.CL
  ),
  position = position_dodge(width = 0.2)) +
  labs(
    title = "Kale",
    x = "Current treatment",
    color = "Historical treatment",
    y = "Exponential growth rate"
  )
```

## Beans

```{r}
formula(b_ht)
```


```{r}
Anova(b_ht)
```

-   days = significant growth (change in height over time)
-   days:hist_treat = different slopes (height over time) for different historical treatments

plot

```{r}
ggplot(my_augment(b_ht), aes(x = days, color = hist_treat, linetype = curr_treat)) + 
  geom_line(aes(y = .fitted)) +
  labs(y = "log(fitted height)")
```

Only significant interaction:

```{r}
my_augment(b_ht) %>% 
  group_by(hist_treat, days) %>% 
  summarize(.fitted = mean(.fitted)) %>% 
ggplot(aes(x = days, color = hist_treat)) + 
  geom_line(aes(y = .fitted)) +
  labs(y = "log(fitted height)")
```

Only slopes:

```{r}
b_slopes <- emtrends(b_ht, ~ hist_treat*curr_treat, var = "days", level = 0.84)
b_slope_df <- 
  b_slopes %>% 
  as_tibble() %>% 
  # back-transform linear slope to exponential growth rate?
  mutate(across(c(days.trend, lower.CL, upper.CL), ~exp(.x)))

ggplot(b_slope_df) +
  geom_pointrange(aes(
    x = curr_treat,
    color = hist_treat,
    y = days.trend,
    ymin = lower.CL,
    ymax = upper.CL
  ),
  position = position_dodge(width = 0.2)) +
  labs(
    title = "Beans",
    x = "Current treatment",
    color = "Historical treatment",
    y = "Exponential growth rate"
  )
```

## Oats

```{r}
formula(o_ht)
```


```{r}
Anova(o_ht)
```

-   days = significant growth (change in height over time)
-   july_herbivory = significant effect of herbivory on height
-   curr_treat = different intercepts for different treatments
-   days:curr_treat = different slopes (height over time) for different current treatments
-   days:hist_treat:curr_treat = hard to interpret, especially since no main effects

```{r}
ggplot(my_augment(o_ht), aes(x = days, color = hist_treat, linetype = curr_treat)) + 
  geom_line(aes(y = .fitted)) +
  labs(y = "predicted height (cm)")
```

Three-way interaction probably driven by low x ambient vs low x 50%

Only current condition interaction:

```{r}
my_augment(o_ht) %>% 
  group_by(curr_treat, days) %>% 
  summarize(.fitted = mean(.fitted)) %>% 
ggplot(aes(x = days, linetype = curr_treat)) + 
  geom_line(aes(y = .fitted)) +
  labs(y = "predicted height (cm)")
```

Just slopes.

```{r}
o_slopes <- emtrends(o_ht, ~ hist_treat*curr_treat, var = "days", level = 0.84)
o_slope_df <- 
  o_slopes %>% 
  as_tibble() %>% 
  # back-transform linear slope to exponential growth rate?
  mutate(across(c(days.trend, lower.CL, upper.CL), ~exp(.x)))

ggplot(o_slope_df) +
  geom_pointrange(aes(
    x = curr_treat,
    color = hist_treat,
    y = days.trend,
    ymin = lower.CL,
    ymax = upper.CL
  ),
  position = position_dodge(width = 0.2)) +
  labs(
    title = "Oats",
    x = "Current treatment",
    color = "Historical treatment",
    y = "Linear growth rate (cm/day)"
  )
```


# References
Payton, M. E., M. H. Greenstone, and N. Schenker. 2003. Overlapping confidence intervals or standard error intervals: What do they mean in terms of statistical significance? Journal of Insect Science 3:1–6.



# Reproducibility

<details>

<summary>

Reproducibility receipt

</summary>

```{r}
## datetime
Sys.time()

## repository
if(requireNamespace('git2r', quietly = TRUE)) {
  git2r::repository()
} else {
  c(
    system2("git", args = c("log", "--name-status", "-1"), stdout = TRUE),
    system2("git", args = c("remote", "-v"), stdout = TRUE)
  )
}

## session info
sessionInfo()
```

</details>
